---
title: "PCA Genotype Analysis"
author: "Anastasiia"
output: 
  html_document:
    toc: TRUE
    df_print: paged
---

```{r setup, include=FALSE}
options(digits = 4, width = 100, stringsAsFactors = T)
knitr::opts_chunk$set(echo = TRUE,
                      tidy.opts = list(width.cutoff = 100),
                      tidy=TRUE,
                      fig.pos = "H",
                      dpi = 600,
                      warning = FALSE, 
                      message = FALSE, 
                      cache = F, 
                      cache.lazy = FALSE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(plyr)
library(data.table)
library(viridis)
```

Before running MDS/PCA, LD pruning was done in PLINK based on pairwise correlation with the following options:

- a window of 50 SNPs: to calculate LD between each pair of SNPs 

- remove one of a pair if the LD is greater than 0.2

- short the window 5 SNPs forward

Unimputed QCed SNP data for the dex samples are located:

```
/binder/common/genotypes/raw_genotypes/genotypes_DexStim/610K/02_Merge/610k_Dex_Batch1_Batch2_QC
```

```
/binder/common/genotypes/raw_genotypes/genotypes_DexStim/OmniExpress/02_Merge/OmniExpress_Dex_Batch1_Batch2_Batch3_Batch4_QC
```


### Merge 610K and Omnni data

Workspace folder with data: ```/binder/mgp/workspace/2020_DexStim_Array_Human/snps/unimputed_qc```

```{sh include = T, eval = F}
cd /binder/mgp/workspace/2020_DexStim_Array_Human/snps/unimputed_qc
plink --bfile 610K/610k_Dex_Batch1_Batch2_QC --bmerge OmniExpress/OmniExpress_Dex_Batch1_Batch2_Batch3_Batch4_QC --make-bed --out merged/610K_OmniExpress_QC
```

Output:

```
Merged fileset written to merged/610K_OmniExpress_QC-merge.bed +
merged/610K_OmniExpress_QC-merge.bim + merged/610K_OmniExpress_QC-merge.fam .
782538 variants loaded from .bim file.
301 people (141 males, 53 females, 107 ambiguous) loaded from .fam.
Ambiguous sex IDs written to merged/610K_OmniExpress_QC.nosex .
26 phenotype values loaded from .fam.
Using 1 thread (no multithreaded calculations invoked).
Before main variant filters, 301 founders and 0 nonfounders present.
Calculating allele frequencies... done.
Warning: 17 het. haploid genotypes present (see merged/610K_OmniExpress_QC.hh
); many commands treat these as missing.
Warning: Nonmissing nonmale Y chromosome genotype(s) present; many commands
treat these as missing.
Total genotyping rate is 0.663868.
782538 variants and 301 people pass filters and QC.
Among remaining phenotypes, 9 are cases and 17 are controls.  (275 phenotypes
are missing.)
--make-bed to merged/610K_OmniExpress_QC.bed + merged/610K_OmniExpress_QC.bim +
merged/610K_OmniExpress_QC.fam ... done.
```

### QC of merged data

```{sh include = T, eval = F}
plink --bfile merged/610K_OmniExpress_QC --geno 0.02 --hwe 1e-3 --maf 0.1  --make-bed —-out qc/dex_geno_unimputed_qc
```

Output:

```
64168 MB RAM detected; reserving 32084 MB for main workspace.
782538 variants loaded from .bim file.
301 people (141 males, 53 females, 107 ambiguous) loaded from .fam.
Ambiguous sex IDs written to qc/dex_geno_unimputed_qc.nosex .
26 phenotype values loaded from .fam.
Using 1 thread (no multithreaded calculations invoked).
Before main variant filters, 301 founders and 0 nonfounders present.
Calculating allele frequencies... done.
Warning: 17 het. haploid genotypes present (see qc/dex_geno_unimputed_qc.hh );
many commands treat these as missing.
Warning: Nonmissing nonmale Y chromosome genotype(s) present; many commands
treat these as missing.
Total genotyping rate is 0.663868.
513297 variants removed due to missing genotype data (--geno).
--hwe: 42 variants removed due to Hardy-Weinberg exact test.
29899 variants removed due to minor allele threshold(s)
(--maf/--max-maf/--mac/--max-mac).
239300 variants and 301 people pass filters and QC.
Among remaining phenotypes, 9 are cases and 17 are controls.  (275 phenotypes
are missing.)
```

### LD Pruning


```{sh include = T, eval = F}
scp -C -r -o ProxyCommand="ssh cluser@biocomp.psych.mpg.de nc slurmgate 22" ~/bio/code/mpip/dex-stim-human-array/data/snps/remove_prune.txt ahryhorzhevska@slurmgate:/binder/mgp/workspace/2020_DexStim_Array_Human/snps/
```

```{sh include = T, eval = F}
plink --bfile qc/dex_geno_unimputed_qc --exclude range ../remove_prune.txt --indep-pairwise 50 5  0.2 --out ld/dex_geno_unimputed_ld
```
Output:

```

64168 MB RAM detected; reserving 32084 MB for main workspace.
239300 variants loaded from .bim file.
301 people (141 males, 53 females, 107 ambiguous) loaded from .fam.
Ambiguous sex IDs written to ld/dex_geno_unimputed_ld.nosex .
26 phenotype values loaded from .fam.
--exclude range: 2304 variants excluded.
--exclude range: 236996 variants remaining.
Using 1 thread (no multithreaded calculations invoked).
Before main variant filters, 301 founders and 0 nonfounders present.
Calculating allele frequencies... done.
Total genotyping rate is 0.999222.
236996 variants and 301 people pass filters and QC.
Among remaining phenotypes, 9 are cases and 17 are controls.  (275 phenotypes
are missing.)
Pruned 13482 variants from chromosome 1, leaving 5502.
Pruned 13972 variants from chromosome 2, leaving 5265.
Pruned 11676 variants from chromosome 3, leaving 4457.
Pruned 9679 variants from chromosome 4, leaving 4084.
Pruned 10384 variants from chromosome 5, leaving 4204.
Pruned 10466 variants from chromosome 6, leaving 3986.
Pruned 9278 variants from chromosome 7, leaving 3683.
Pruned 9624 variants from chromosome 8, leaving 3342.
Pruned 8764 variants from chromosome 9, leaving 3309.
Pruned 9235 variants from chromosome 10, leaving 3633.
Pruned 8941 variants from chromosome 11, leaving 3314.
Pruned 8484 variants from chromosome 12, leaving 3434.
Pruned 6265 variants from chromosome 13, leaving 2531.
Pruned 5666 variants from chromosome 14, leaving 2303.
Pruned 5162 variants from chromosome 15, leaving 2201.
Pruned 5300 variants from chromosome 16, leaving 2409.
Pruned 4494 variants from chromosome 17, leaving 2312.
Pruned 5081 variants from chromosome 18, leaving 2280.
Pruned 3048 variants from chromosome 19, leaving 1803.
Pruned 4339 variants from chromosome 20, leaving 2039.
Pruned 2666 variants from chromosome 21, leaving 1155.
Pruned 2507 variants from chromosome 22, leaving 1237.
Pruning complete.  168513 of 236996 variants removed.
Marker lists written to ld/dex_geno_unimputed_ld.prune.in and
ld/dex_geno_unimputed_ld.prune.out .
```

```{sh include = T, eval = F}
plink --bfile qc/dex_geno_unimputed_qc --extract ld/dex_geno_unimputed_ld.prune.in --genome --min 0.0625 --make-bed --out pca/ld_extracted/dex_geno_unimputed_ld
```

Output:
```
64168 MB RAM detected; reserving 32084 MB for main workspace.
239300 variants loaded from .bim file.
301 people (141 males, 53 females, 107 ambiguous) loaded from .fam.
Ambiguous sex IDs written to pca/ld_extracted/dex_geno_unimputed_ld.nosex .
26 phenotype values loaded from .fam.
--extract: 68483 variants remaining.
Using up to 31 threads (change this with --threads).
Before main variant filters, 301 founders and 0 nonfounders present.
Calculating allele frequencies... done.
Total genotyping rate is 0.999235.
68483 variants and 301 people pass filters and QC.
Among remaining phenotypes, 9 are cases and 17 are controls.  (275 phenotypes
are missing.)
--make-bed to pca/ld_extracted/dex_geno_unimputed_ld.bed +
pca/ld_extracted/dex_geno_unimputed_ld.bim +
pca/ld_extracted/dex_geno_unimputed_ld.fam ... done.
IBD calculations complete.
Finished writing pca/ld_extracted/dex_geno_unimputed_ld.genome .
```

### MDS with plink

<!-- ```{sh include = T, eval = F} -->
<!-- plink --bfile qc/dex_geno_unimputed_qc --extract ld/dex_geno_unimputed_ld.prune.in --read-genome pca/ld_extracted/dex_geno_unimputed_ld  --cluster --mds-plot 20 'eigvals' 'eigendecomp' --pca --out pca/mds/dex_geno_unimputed_ldpruned_mds -->
<!-- ``` -->

```{sh include = T, eval = F}
plink --bfile pca/ld_extracted/dex_geno_unimputed_ld  --cluster --mds-plot 20 'eigvals' 'eigendecomp' --pca --out pca/mds/dex_geno_unimputed_ld_mds
```

Output:

```
64168 MB RAM detected; reserving 32084 MB for main workspace.
68483 variants loaded from .bim file.
301 people (141 males, 53 females, 107 ambiguous) loaded from .fam.
Ambiguous sex IDs written to pca/mds/dex_geno_unimputed_ld_mds.nosex .
26 phenotype values loaded from .fam.
Using up to 31 threads (change this with --threads).
Before main variant filters, 301 founders and 0 nonfounders present.
Calculating allele frequencies... done.
Total genotyping rate is 0.999235.
68483 variants and 301 people pass filters and QC.
Among remaining phenotypes, 9 are cases and 17 are controls.  (275 phenotypes
are missing.)
Relationship matrix calculation complete.
--pca: Results saved to pca/mds/dex_geno_unimputed_ld_mds.eigenval and
pca/mds/dex_geno_unimputed_ld_mds.eigenvec .
Distance matrix calculation complete.
Clustering... done.
Cluster solution written to pca/mds/dex_geno_unimputed_ld_mds.cluster1 ,
pca/mds/dex_geno_unimputed_ld_mds.cluster2 , and
pca/mds/dex_geno_unimputed_ld_mds.cluster3 .
Performing multidimensional scaling analysis (eigendecomposition algorithm,
20 dimensions)... done.
MDS solution written to pca/mds/dex_geno_unimputed_ld_mds.mds (eigenvalues in
pca/mds/dex_geno_unimputed_ld_mds.mds.eigvals ).
```

### MDS analysis

```{sh include = T, eval = F}
cd ~/bio/workspace/dex/dex-genotype/pca/
scp -C -r -o ProxyCommand="ssh cluser@biocomp.psych.mpg.de nc slurmgate 22" ahryhorzhevska@slurmgate:/binder/mgp/workspace/2020_DexStim_Array_Human/snps/unimputed_qc/ .
```

```{r}
eigenvec.fn  <- "~/bio/workspace/dex/dex-genotype/pca/unimputed_qc/pca/mds/dex_geno_unimputed_ld_mds.mds"
eigenval.fn  <- "~/bio/workspace/dex/dex-genotype/pca/unimputed_qc/pca/mds/dex_geno_unimputed_ld_mds.mds.eigvals"

eigenvec     <- fread(eigenvec.fn)
eigenval     <- fread(eigenval.fn)
```

#### Eigenvalues: % of variance

```{r}
pve <- data.frame(PC = 1:nrow(eigenval), pve = eigenval/sum(eigenval) * 100)
names(pve)[2] <- "pve" 

ggplot(pve[1:30,], aes(PC, pve)) + 
  geom_bar(stat = "identity") + 
  ylab("Percentage variance explained") + 
  theme( panel.background = element_blank(),
         plot.title = element_text(size = 10),
         axis.title = element_text(size = 10),
         axis.text.x = element_text(angle = 0, hjust = 1), 
         legend.position = "none")
```

#### Take only individuals who are in the Dex Stim Study

```{r}
pheno.methyl.fn <- "/Users/anastasiia_hry/bio/datasets/pheno/pheno_full_for_kimono.csv"
pheno           <- fread(pheno.methyl.fn)

pheno           <- pheno[!is.na(DNAm_ID),]
# eigenval        <- eigenval[eigenvec$FID %in% pheno$DNA_ID, ]
eigenvec        <- eigenvec[eigenvec$FID %in% pheno$DNA_ID, ]
```


#### MDS Individual plot

```{r}
ggplot(data = eigenvec, aes(C1, C2, col = as.factor(SOL))) + 
  geom_vline(xintercept = 0, linetype = "dashed", size = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 0.4) +
  geom_point(size = 1, alpha = 1) +
  xlab(paste0("PC1 (", signif(pve$pve[1], 2), "%)")) + 
  ylab(paste0("PC2 (", signif(pve$pve[2], 2), "%)")) +
  labs(col = "Land") +
  theme( panel.background = element_blank(),
         plot.title = element_text(size = 10),
         axis.title = element_text(size = 10),
         axis.text.x = element_text(angle = 0, hjust = 1), 
         legend.position = "none")
```

#### K-means clustering

```{r}
kmeans.mdl <- kmeans(eigenvec[, -c(1:3)], centers = 2, nstart = 10)            
clusters   <- kmeans.mdl$cluster
```


```{r include=F}
pca        <- cbind(eigenvec,  clusters)
ggplot(data = pca, aes(C1, C2, col = as.factor(clusters))) + 
  geom_point() +
  theme(legend.position = "none") + 
 # xlab(paste0("PC1 (", signif(pve$pve[1], 2), "%)")) + 
 #  ylab(paste0("PC2 (", signif(pve$pve[2], 2), "%)")) +
  theme( panel.background = element_blank(),
         plot.title = element_text(size = 10),
         axis.title = element_text(size = 10),
         axis.text.x = element_text(angle = 0, hjust = 1), 
         legend.position = "none")
  
```
#### Hierarhical clustering

```{r}
d      <- dist(eigenvec[, -c(1:3)], method = "euclidean")
hc.mdl <- hclust(d, method = "ward" )
plot(hc.mdl)
```

```{r}
sub_grp <- as.factor(cutree(hc.mdl, k = 2)) # length(kmeans.mdl$size))
eigenvec <- left_join(eigenvec, pheno[Dex == 1, .(DNA_ID, Status, SNP_Chip)], by = c("FID" = "DNA_ID"))
ggplot(data = eigenvec, aes(C1, C2, col = sub_grp)) + # as.factor(SNP_Chip))) + # as.factor(Status))) +  
  geom_vline(xintercept = 0, linetype = "dashed", size = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 0.4) +
  geom_point(size = 2, alpha = 1) +
 # geom_text(aes(label = FID), hjust = 0.8, nudge_y = 0.003) + 
  # scale_color_brewer(palette = "Paired", na.value = "light grey") +
  # scale_color_viridis(discrete = TRUE, option = "plasma", na.value = "light grey") +
  xlab(paste0("PC1 (", signif(pve$pve[1], 2), "%)")) + 
  ylab(paste0("PC2 (", signif(pve$pve[2], 2), "%)")) +
  theme( panel.background = element_blank(),
         plot.title = element_text(size = 10),
         axis.title = element_text(size = 10),
         axis.text.x = element_text(angle = 0, hjust = 1), 
         legend.position = "bottom")
```

#### Calculate mean distance of each sample from teh mean of PC1 and OC2

```{r}
eps     <- 0.01
pc.mean <- mean(eigenvec$C1)
pc.std  <- sd(eigenvec$C1)
cut.off <- 4 * pc.std
lower   <- pc.mean - cut.off + eps
upper   <- pc.mean + cut.off - eps

outliers <- eigenvec[C1 < lower | C1 > upper, FID ]
outliers
```


#### Add etnicity information


```{r}
library(haven)
etnicity.fn <- "/Users/anastasiia_hry/bio/datasets/pheno/MAD_GSK_Phenotype_20191009_for_ah_02072021.sav"
etnicity    <- read_sav(etnicity.fn)

etnicity    <- etnicity[etnicity$NID %in% pheno$DNA_ID, ]
pheno       <- left_join(pheno, etnicity[, c("NID", "S_Land")], by = c("DNA_ID" = "NID"))
eigenvec    <- left_join(eigenvec, etnicity[, c("NID", "S_Land")], by = c("FID" = "NID"))

na.id <- pheno[!(DNA_ID %in% etnicity$NID), "DNA_ID"]
na.id <- na.id[!duplicated(na.id$DNA_ID), ]
```

```{r}
ggplot(data = eigenvec, aes(C1, C2, color = S_Land)) +  
  geom_vline(xintercept = 0, linetype = "dashed", size = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 0.4) +
  geom_point(size = 1, alpha = 1) +
  xlab(paste0("PC1 (", signif(pve$pve[1], 2), "%)")) + 
  ylab(paste0("PC2 (", signif(pve$pve[2], 2), "%)")) +
  labs(col = "Land") +
  theme( panel.background = element_blank(),
         plot.title = element_text(size = 10),
         axis.title = element_text(size = 10),
         axis.text.x = element_text(angle = 0, hjust = 1), 
         legend.position = "right")
```
