---
title: "Cell-type specific enrichment analysis"
subtitle: Dexamethasone-Stimulated Human Array Project
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---

```{r, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(corrplot)
library(factoextra)
library(viridis)
library(ggpubr)
library(UpSetR)
```

```{r}
cbPalette <- c( "#0072B2", "#009E73", "#E69F00", "#F0E442", "#D55E00", "#CC79A7", "#56B4E9", "#999999")
separate <- ""
```

```{r}
treatment   <- "veh"

pval.veh.df <- fread(paste0("~/bio/code/mpip/dex-stim-human-array/output/data/integrative/cell_type_enrichment/dnam_cell_type_enrichment_pvals_",treatment, separate, ".csv"))

pval.veh.bcc.df <- pval.veh.df[, 1:13]
```

```{r}
fdr.bcc.df <- matrix(p.adjust(as.vector(as.matrix(pval.veh.bcc.df[, 2:13])), method='fdr'), 
                     ncol=12) %>%
  data.frame()

fdr.bcc.df <- cbind(pval.veh.bcc.df$CpG_ID, fdr.bcc.df)
colnames(fdr.bcc.df) <- colnames(pval.veh.bcc.df)
```

```{r}
sign.pval.df <- fdr.bcc.df %>% reshape2::melt(measure.vars = colnames(fdr.bcc.df)[2:13]) %>% setDT()
colnames(sign.pval.df) <- c("CpG_ID", "Type", "fdr")

sign.pval.df <- sign.pval.df[fdr <= 0.05]
sign.pval.df <- na.omit(sign.pval.df)
sign.pval.df
```

```{r}
veh.sign.pval.df <- sign.pval.df
veh.sign.pval.df[["Treatment"]] <- treatment
```

DEX

```{r}
treatment   <- "dex"

pval.dex.df <- fread(paste0("~/bio/code/mpip/dex-stim-human-array/output/data/integrative/cell_type_enrichment/dnam_cell_type_enrichment_pvals_",treatment, separate, ".csv"))

pval.dex.bcc.df <- pval.dex.df[, 1:13]
```

```{r}
fdr.bcc.df <- matrix(p.adjust(as.vector(as.matrix(pval.dex.bcc.df[, 2:13])), method ='fdr'), 
                     ncol=12) %>%
  data.frame()

fdr.bcc.df <- cbind(pval.dex.bcc.df$CpG_ID, fdr.bcc.df)
colnames(fdr.bcc.df) <- colnames(pval.dex.bcc.df)
```

```{r}
sign.pval.df <- fdr.bcc.df %>% reshape2::melt(measure.vars = colnames(fdr.bcc.df)[2:13]) %>% setDT()
colnames(sign.pval.df) <- c("CpG_ID", "Type", "fdr")
sign.pval.df <- sign.pval.df[fdr <= 0.05]

sign.pval.df <- na.omit(sign.pval.df)
sign.pval.df
```

```{r}
dex.sign.pval.df <- sign.pval.df
dex.sign.pval.df[["Treatment"]] <- treatment
```

## Distribution plots for Basleine and Dex together

```{r}
sign.pval.df <- rbind(veh.sign.pval.df, dex.sign.pval.df)
intersect.cpgs <- intersect(dex.sign.pval.df$CpG_ID, veh.sign.pval.df$CpG_ID)

print(paste0("Number of unique dex CpGs: ", length(unique(dex.sign.pval.df$CpG_ID))))
print(paste0("Number of unique baseline CpGs: ", length(unique(veh.sign.pval.df$CpG_ID))))
print(paste0("Number of intesecting CpGs: ", length(unique(intersect.cpgs))))
```

```{r}
ggplot(sign.pval.df[CpG_ID %in% intersect.cpgs], aes(x = Type, fill = Treatment)) +
  geom_bar(stat = "count", alpha = 1, position = position_dodge(width = .9)) +
  # geom_text(aes(label =  scales::percent(prop.table(stat(count)))), 
  # geom_text(aes(label = scales::percent(stat(count) / length(unique(sign.pval.df$CpG_ID)))), 
  geom_text(aes(label = scales::percent(stat(count) / length(intersect.cpgs), accuracy = 0.1)), 
            stat = "count", vjust = -0.5, size = 5, position = position_dodge(width = .9)) + 
  theme(legend.position = "bottom", # c(0.9, 0.9), 
        legend.title = element_blank(), 
        legend.text = element_text(size = 18),
        panel.grid.major = element_blank(), 
        panel.background = element_blank(),
        plot.title = element_text(size = 18), 
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 18)) + 
  labs(title = "Distribution of CpGs significant at FDR = 0.05 across 12 blood cell types",
       x = "", y = "") +
  scale_fill_manual(values = cbPalette[2:3]) 
```

```{r eval = F, include = F}
ggplot(sign.pval.df, aes(x = Type, group = Treatment, fill = Treatment)) + 
  geom_density(aes(y = stat(count) / nrow(fdr.bcc.df), color = Treatment), 
           stat = "count",
           alpha = 0.2) + 
  scale_y_continuous(labels = scales::percent) + 
  labs(x = " ", y = "", 
       title = "Distribution of CpGs significant at FDR = 0.05 across 12 blood cell types") + 
  theme(legend.title = element_blank(), 
        legend.position = "bottom", 
        panel.grid.major = element_blank(), 
        panel.background = element_blank(), 
        plot.title = element_text(size = 8), 
        axis.title = element_text(size = 8), 
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  scale_color_manual(values = cbPalette[2:3]) +
  scale_fill_manual(values = cbPalette[2:3]) 
```

# Cell-type specificity on baseline-meQTLs

```{r eval = F}
meqtl.veh.df          <- fread("~/bio/code/mpip/dex-stim-human-array/output/data/integrative/matrixEQTL/meqtls/me-qtl_cis_result_veh_fdr_005.csv",
                           col.names = c("SNP", "CpG_ID", "beta", "t-stat", "p-value", "FDR")) 
veh.meqtl.cpgs        <- meqtl.veh.df$CpG_ID %>% unique() 
sign.pval.veh.mqtl.df <- sign.pval.df[CpG_ID %in% veh.meqtl.cpgs, ] 

sign.pval.cpgs.dex    <- sign.pval.veh.mqtl.df[Treatment == "dex", CpG_ID]
sign.pval.cpgs.veh    <- sign.pval.veh.mqtl.df[Treatment == "veh", CpG_ID]

intersect.cpgs        <- intersect(sign.pval.cpgs.dex, sign.pval.cpgs.veh) 
sign.pval.veh.mqtl.df <- sign.pval.veh.mqtl.df[CpG_ID %in% intersect.cpgs,]

print(paste0("Number of unique dex CpGs: ", length(unique(sign.pval.cpgs.dex))))
print(paste0("Number of unique baseline CpGs: ", length(unique(sign.pval.cpgs.veh))))
print(paste0("Number of intesecting CpGs: ", length(unique(intersect.cpgs))))

blood.types.df.lst  <- split(sign.pval.veh.mqtl.df[Treatment == "veh"], by = "Type")
blood.types.cpg.lst <- lapply(blood.types.df.lst, function(df) df$CpG_ID)

saveRDS(blood.types.cpg.lst, "~/bio/code/mpip/dex-stim-human-array/output/data/integrative/cell_type_enrichment/veh_meqtl_cpgs_by_blood_cell_type_significant.rds")

# SNPs

sign.pval.veh.mqtl.with.snps.df <- left_join(sign.pval.veh.mqtl.df, meqtl.veh.df[,.(SNP, CpG_ID)])

blood.types.df.lst  <- split(sign.pval.veh.mqtl.with.snps.df[sign.pval.veh.mqtl.with.snps.df$Treatment == "veh",], by = "Type")
blood.types.snp.lst <- lapply(blood.types.df.lst, function(df) df$SNP)

saveRDS(blood.types.snp.lst, "~/bio/code/mpip/dex-stim-human-array/output/data/integrative/cell_type_enrichment/veh_meqtl_snps_by_blood_cell_type_significant.rds")

saveRDS(blood.types.df.lst, "~/bio/code/mpip/dex-stim-human-array/output/data/integrative/cell_type_enrichment/veh_meqtls_by_blood_cell_type_significant.rds")
```

# Cell-type specificity on GR-induced (delta)-meQTLs

```{r}
ind.meqtl.delta.df      <- fread("~/bio/code/mpip/dex-stim-human-array/output/data/integrative/matrixEQTL/meqtls/region_wise_independent_snps/me-qtl_cis_indp_rw_delta_fdr_005.csv") 
delta.meqtl.cpgs        <- ind.meqtl.delta.df$CpG_ID %>% unique() 
sign.pval.delta.mqtl.df <- sign.pval.df[CpG_ID %in% delta.meqtl.cpgs, ] 

sign.pval.cpgs.dex <- sign.pval.delta.mqtl.df[Treatment == "dex", CpG_ID]
sign.pval.cpgs.veh <- sign.pval.delta.mqtl.df[Treatment == "veh", CpG_ID]

intersect.cpgs          <- intersect(sign.pval.cpgs.dex, sign.pval.cpgs.veh) 
sign.pval.delta.mqtl.df <- sign.pval.delta.mqtl.df[CpG_ID %in% intersect.cpgs,]

print(paste0("Number of unique dex CpGs: ", length(unique(sign.pval.cpgs.dex))))
print(paste0("Number of unique baseline CpGs: ", length(unique(sign.pval.cpgs.veh))))
print(paste0("Number of intesecting CpGs: ", length(unique(intersect.cpgs))))
```


## Check if changes are significant

### Enrichment GR-meQTL CpGs over all CpGs

```{r eval = F, include = F}
blood.cell.types        <- unique(sign.pval.delta.mqtl.df$Type)

enrich.res.df <- lapply(blood.cell.types, function(i){  # for each blood cell type
  
  ### DEX ###
  
  treatment <- "dex"
  
  type.sign      <- nrow(sign.pval.df[Type == i][Treatment == treatment])
  type.non.sign  <- nrow(sign.pval.df[Treatment == treatment]) - type.sign
  meqtl.sign     <- nrow(sign.pval.delta.mqtl.df[Type == i][Treatment == treatment])
  meqtl.non.sign <- nrow(sign.pval.delta.mqtl.df[Treatment == treatment]) - meqtl.sign
  
  conf.mtrx <- matrix(c(type.sign, meqtl.sign, 
                        type.non.sign, meqtl.non.sign), 
                      2, 2, byrow = TRUE)
  
  test.rslt <- fisher.test(conf.mtrx)
  dex.df    <- data.frame(Type = i, p.value = test.rslt$p.value, Treatment = treatment)
  
  ### Basleine ###
  
  treatment <- "veh"
  
  type.sign      <- nrow(sign.pval.df[Type == i][Treatment == treatment])
  type.non.sign  <- nrow(sign.pval.df[Treatment == treatment]) - type.sign
  meqtl.sign     <- nrow(sign.pval.delta.mqtl.df[Type == i][Treatment == treatment])
  meqtl.non.sign <- nrow(sign.pval.delta.mqtl.df[Treatment == treatment]) - meqtl.sign
  
  conf.mtrx <- matrix(c(type.sign, meqtl.sign, 
                        type.non.sign, meqtl.non.sign), 
                      2, 2, byrow = TRUE)
  
  test.rslt <- fisher.test(conf.mtrx)
  veh.df    <- data.frame(Type = i, p.value = test.rslt$p.value, Treatment = treatment)
  
  return(rbind(dex.df, veh.df))
  }) %>%
  bind_rows()

reshape(enrich.res.df, idvar = "Type", timevar = "Treatment", direction = "wide")
```

```{r eval = F, include = F}
ggplot(sign.pval.delta.mqtl.df, aes(x = Type, fill = Treatment)) +
  geom_bar(stat = "count", alpha = 1, position = position_dodge(width = .9)) +
  geom_text(aes(label = scales::percent(stat(count) / length(intersect.cpgs), accuracy = 1)), 
            stat = "count", vjust = 1.5, size = 5, position = position_dodge(width = .9)) + 
  geom_label(data = subset(enrich.res.df, p.value <= 0.05),
            aes(label = "*"), 
            stat = "count", vjust = -1.5, size = 5, position = position_dodge(width = 1)) +
  theme(legend.position = "bottom", # c(0.9, 0.9), 
        legend.title = element_blank(), 
        legend.text = element_text(size = 18),
        panel.grid.major = element_blank(), 
        panel.background = element_blank(),
        plot.title = element_text(size = 18), 
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 18, colour = "black")) + 
  labs(title = "Distribution of CpGs significant at FDR = 0.05 across 12 blood cell types",
       x = "", y = "") +
  scale_fill_manual(values = cbPalette[2:3]) 
```

### Distribution of P-values for each blood cell-type

Kolmogorov-Smirnov test: Do 2 samples follow the same distribution?

```{r}
blood.cell.types        <- unique(sign.pval.delta.mqtl.df$Type)
# blood.cell.types        <- levels(sign.pval.delta.mqtl.df$Type)

test.res.df <- lapply(blood.cell.types, function(i){  # for each blood cell type
  dex.fdr.lst <- sign.pval.delta.mqtl.df[Type == i][Treatment == "dex", fdr]
  veh.fdr.lst <- sign.pval.delta.mqtl.df[Type == i][Treatment == "veh", fdr]
  test.rslt   <- ks.test(dex.fdr.lst, veh.fdr.lst)
 
  return(data.frame(Type = i, "p-value" = test.rslt$p.value))
  }) %>%
  bind_rows()

sign.pval.delta.mqtl.df <- left_join(sign.pval.delta.mqtl.df, test.res.df)
```

```{r}
ggplot(sign.pval.delta.mqtl.df, aes(x = -log10(fdr), fill = Treatment)) + 
  geom_density(alpha = 0.3, aes(color = Treatment)) +
  geom_text(data = subset(unique(sign.pval.delta.mqtl.df[, .(Treatment, Type, p.value)])),
            aes(x = -Inf, y = -Inf, label = paste0("KS test = ", signif(p.value, 3))), 
            hjust = -0.5, vjust = -15, size = 5, color = "black" ) +
  facet_wrap(~ Type, scales = "free", ncol = 4) +
  theme(legend.position = "bottom", # c(.9,.9), 
        legend.title = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(size = 10, color = "black"),
        axis.title = element_text(size = 10, color = "black"),
        axis.text.x = element_text(size = 10, color = "black"),
        axis.text.y = element_text(size = 10, color = "black"),
        strip.text.x = element_text(size = 10, margin = margin())) +
  labs(title = "Distribution of FDRs of GR-meQTL CpGs across bloss cell type", x = "-log10 FDR", y = "") +
  scale_fill_manual(values = cbPalette[2:3]) +
  scale_color_manual(values = cbPalette[2:3]) 
```

```{r}
plt.size <- 12
cbPalette <-c( "grey", "#0072B2")
sign.pval.delta.mqtl.df <- sign.pval.delta.mqtl.df %>% setDT()
sign.pval.delta.mqtl.df[Treatment == "veh", Treatment := "Baseline"]
sign.pval.delta.mqtl.df[Treatment == "dex", Treatmenet := "GR-response"]

ggplot(sign.pval.delta.mqtl.df, aes(x = Type, fill = Treatment)) +
  geom_bar(stat = "count", alpha = 1, position = position_dodge(width = .9)) +
  geom_text(aes(label = scales::percent(stat(count) / length(intersect.cpgs), accuracy = 0.1)), 
            stat = "count", vjust = -0.5, size = plt.size / 3, position = position_dodge(width = 0.9)) + 
  geom_text(data = subset(sign.pval.delta.mqtl.df[Treatment == "Baseline"], p.value <= 0.05),
            aes(label = "*"), stat = "count", vjust = 0.05, hjust = 0., size = plt.size/2 + 2) +
  theme(legend.position = "bottom", # c(0.9, 0.9), 
        legend.title = element_blank(), 
        legend.text = element_text(size = plt.size),
        panel.grid.major = element_blank(), 
        panel.background = element_blank(),
        plot.title = element_text(size = plt.size), 
        axis.title = element_text(size = plt.size),
        axis.text = element_text(size = plt.size, colour = "black")) + 
  labs(title = "Distribution of CpGs significant at FDR = 0.05 across 12 blood cell types",
       x = "", y = "") +
  scale_fill_manual(values = cbPalette) 
```

## Upset plot for DEX CpGs

```{r}
blood.types.df.lst  <- split(sign.pval.delta.mqtl.df[Treatment == "dex"], by = "Type")
blood.types.cpg.lst <- lapply(blood.types.df.lst, function(df) df$CpG_ID)

saveRDS(blood.types.cpg.lst, "~/bio/code/mpip/dex-stim-human-array/output/data/integrative/cell_type_enrichment/dex_meqtl_cpgs_by_blood_cell_type_significant.rds")

upset(fromList(blood.types.cpg.lst), 
      nsets = 12, 
      nintersects = 75, 
      queries = list(list(query = intersects, params = list("Bas"), color = "red"),
                     list(query = intersects, params = list("Bmem"), color = "red"),
                     list(query = intersects, params = list("Bnv"), color = "red"),
                     list(query = intersects, params = list("CD4mem"), color = "red"),
                     list(query = intersects, params = list("CD4nv"), color = "red"),
                     list(query = intersects, params = list("CD8mem"), color = "red"),
                     list(query = intersects, params = list("Neu"), color = "red")),
      mainbar.y.label = "Number of intersections \ndex CpGs", 
      keep.order = T,
      order.by = "freq", 
      text.scale = 1.3, 
      line.size = 0.1,
      mb.ratio = c(0.4, 0.6))
```

## Upset plot for DEX SNPs

```{r}
sign.pval.delta.mqtl.with.snps.df <- left_join(sign.pval.delta.mqtl.df, ind.meqtl.delta.df[,.(SNP, CpG_ID)])

blood.types.df.lst  <- split(sign.pval.delta.mqtl.with.snps.df[sign.pval.delta.mqtl.with.snps.df$Treatment == "dex",], by = "Type")
blood.types.snp.lst <- lapply(blood.types.df.lst, function(df) unique(df$SNP))

# saveRDS(blood.types.df.lst, "~/bio/code/mpip/dex-stim-human-array/output/data/integrative/cell_type_enrichment/delta_meqtls_by_blood_cell_type_significant.rds")

# saveRDS(blood.types.snp.lst, "~/bio/code/mpip/dex-stim-human-array/output/data/integrative/cell_type_enrichment/dex_meqtl_snps_by_blood_cell_type_significant.rds")

upset(fromList(blood.types.snp.lst), 
      nsets = 12, 
      nintersects = 75, 
      queries = list(list(query = intersects, params = list("Bas"), color = "red"),
                     list(query = intersects, params = list("Bmem"), color = "red"),
                     list(query = intersects, params = list("Bnv"), color = "red"),
                     list(query = intersects, params = list("CD4mem"), color = "red"),
                     list(query = intersects, params = list("CD4nv"), color = "red"),
                     list(query = intersects, params = list("CD8mem"), color = "red"),
                     list(query = intersects, params = list("Neu"), color = "red")),
      mainbar.y.label = "Number of intersections \ndex SNPs", 
      keep.order = T,
      order.by = "freq", 
      text.scale = 1.3, 
      line.size = 0.1,
      mb.ratio = c(0.4, 0.6))
```

# Cell-type specificity on DMPs

```{r}
dmps.sign.anno.fn <- "~/bio/code/mpip/dex-stim-human-array/output/data/methylation/02_dmp/dmps_fdr01_fc02_anno_full.csv"
dmps.sign.anno.df <- fread(dmps.sign.anno.fn)
dmps              <- dmps.sign.anno.df$PROBE_ID %>% unique() 
sign.pval.dmps.df <- sign.pval.df[CpG_ID %in% dmps, ] 

sign.pval.dmps.dex <- sign.pval.dmps.df[Treatment == "dex", CpG_ID]
sign.pval.dmps.veh <- sign.pval.dmps.df[Treatment == "veh", CpG_ID]

intersect.dmps     <- intersect(sign.pval.dmps.dex, sign.pval.dmps.veh) 
sign.pval.dmps.df  <- sign.pval.dmps.df[CpG_ID %in% intersect.dmps,]

print(paste0("Number of unique dex CpGs: ", length(unique(sign.pval.dmps.dex))))
print(paste0("Number of unique baseline CpGs: ", length(unique(sign.pval.dmps.veh))))
print(paste0("Number of intesecting CpGs: ", length(unique(intersect.dmps))))
```

Kolmogorov-Smirnov test: Do 2 samples follow the same distribution?

```{r}
blood.cell.types        <- unique(sign.pval.delta.mqtl.df$Type)

test.res.df <- lapply(blood.cell.types, function(i){  # for each blood cell type
  dex.fdr.lst <- sign.pval.dmps.df[Type == i][Treatment == "dex", fdr]
  veh.fdr.lst <- sign.pval.dmps.df[Type == i][Treatment == "veh", fdr]
  test.rslt   <- ks.test(dex.fdr.lst, veh.fdr.lst)
 
  return(data.frame(Type = i, "p-value" = test.rslt$p.value))
  }) %>%
  bind_rows()

sign.pval.dmps.df <- left_join(sign.pval.dmps.df, test.res.df)
```

```{r}
ggplot(sign.pval.dmps.df, aes(x = -log10(fdr), fill = Treatment)) + 
  geom_density(alpha = 0.3, aes(color = Treatment)) +
  geom_text(data = subset(unique(sign.pval.dmps.df[, .(Treatment, Type, p.value)])),
            aes(x = -Inf, y = -Inf, label = paste0("KS test = ", signif(p.value, 3))), 
            hjust = -0.5, vjust = -15, size = 5, color = "black" ) +
  facet_wrap(~ Type, scales = "free", ncol = 4) +
  theme(legend.position = "bottom", # c(.9,.9), 
        legend.title = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(size = 10, color = "black"),
        axis.title = element_text(size = 10, color = "black"),
        axis.text.x = element_text(size = 10, color = "black"),
        axis.text.y = element_text(size = 10, color = "black"),
        strip.text.x = element_text(size = 10, margin = margin())) +
  labs(title = "Distribution of FDRs of GR-DMPs across bloss cell type", x = "-log10 FDR", y = "") +
  scale_fill_manual(values = cbPalette[2:3]) +
  scale_color_manual(values = cbPalette[2:3]) 
```

```{r}
ggplot(sign.pval.dmps.df, aes(x = Type, fill = Treatment)) +
  geom_bar(stat = "count", alpha = 1, position = position_dodge(width = .9)) +
  geom_text(aes(label = scales::percent(stat(count) / length(intersect.dmps), accuracy = 0.1)), 
            stat = "count", vjust = -0.1, size = 5, position = position_dodge(width = .9)) + 
  geom_text(data = subset(sign.pval.dmps.df[Treatment == "veh"], p.value <= 0.05),
             aes(label = "*"), stat = "count", vjust = 0.15,size = 16) +
  theme(legend.position = "bottom", # c(0.9, 0.9), 
        legend.title = element_blank(), 
        legend.text = element_text(size = 16),
        panel.grid.major = element_blank(), 
        panel.background = element_blank(),
        plot.title = element_text(size = 16), 
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 16, colour = "black")) + 
  labs(title = "Distribution of GR-DMPs across 12 blood cell types",
       x = "", y = "") +
  scale_fill_manual(values = cbPalette[2:3]) 
```
